---
title: "Exploratory Data Analysis"
output:
  html_document:
    df_print: paged
---
# Import Essential Libraries
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
library(DescTools)
```

# Univariate EDA
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Read in Approved Loan Data
df_approved = read_csv("E:/josh_data_science/Lending Club Loan Data/approved_loan_data_ver3.csv")
```

### Requested Loan Amount
```{r}
Desc(df_approved['loan_amnt'], main = "Loan amount distribution for Approved Loans", plotit = TRUE)
```

### Grade/Subgrade
```{r}
ggplot(df_approved,aes(x=sub_grade)) + geom_bar(stat='count')
```

### Loan Status

```{r}
Desc(df_approved$loan_status, main= "Loan Status Distribution in approved loans", plotit = T)
```

### Debt to Income Ratio (DTI)
```{r}
Desc(df_approved['dti'], main = "Debt to Income Ratio distribution for Approved Loans", plotit = TRUE)
```

### home_ownership
```{r}
df_approved %>% group_by(home_ownership) %>% dplyr::summarise(count=n()) %>% 
mutate(pct=count*100/sum(count))%>% 
  ggplot(aes(x = reorder(home_ownership, -pct), y = pct)) + geom_bar(stat = "identity", fill = 'lightblue',aes(color = I('black')), size = 0.1) + 
xlab("Home Ownership") + ylab("Percent")
```

### Annual_inc
```{r}
Desc(df_approved['annual_inc'], main = "Annual Income Distribution of Approved Loans", plotit = TRUE)
```

### Employment Length
```{r}
ggplot(df_approved,aes(x=emp_length,fill=emp_length))+geom_bar(stat="count")+labs(x="Employment Length",title="Distribution of Employment Length")
```

### Loan Descriptions
```{r}
charged_off_df = df_approved %>% filter(loan_status=='Charged Off')
late_df = df_approved %>% filter(loan_status=='Late (16-30 days)' | loan_status=='Late (31-120 days)')
fully_paid_df = df_approved %>% filter(loan_status == 'Fully Paid')
```

```{r}
default_df = df_approved %>% filter(loan_status == 'Default')
```

```{r}
library(tm)
library(RColorBrewer)
library(wordcloud)

# WordCloud of loan title of Charged Off Loans

loan_descriptions.corpus <- Corpus(VectorSource(charged_off_df['title']))
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removePunctuation)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removeNumbers)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, tolower)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removeWords, stopwords("english"))

wordcloud(loan_descriptions.corpus,
          max.words = 100,
          random.order=FALSE, 
          rot.per=0.30, 
          use.r.layout=FALSE, 
          colors=brewer.pal(8, "Paired"))
```

```{r}
# WordCloud of loan title of Late Paid Loans
loan_descriptions.corpus <- Corpus(VectorSource(late_df['title']))
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removePunctuation)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removeNumbers)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, tolower)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removeWords, stopwords("english"))

wordcloud(loan_descriptions.corpus,
          max.words = 100,
          random.order=FALSE, 
          rot.per=0.30, 
          use.r.layout=FALSE, 
          colors=brewer.pal(8, "Paired"))
```
```{r}
# WordCloud of loan title of Defaulted Loans
loan_descriptions.corpus <- Corpus(VectorSource(default_df['title']))
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removePunctuation)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removeNumbers)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, tolower)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removeWords, stopwords("english"))

wordcloud(loan_descriptions.corpus,
          max.words = 100,
          random.order=FALSE, 
          rot.per=0.30, 
          use.r.layout=FALSE, 
          colors=brewer.pal(8, "Paired"))
```
```{r}
# WordCloud of loan title of Fully Paid Loans
loan_descriptions.corpus <- Corpus(VectorSource(fully_paid_df['title']))
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removePunctuation)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removeNumbers)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, tolower)
loan_descriptions.corpus <- tm_map(loan_descriptions.corpus, removeWords, stopwords("english"))

wordcloud(loan_descriptions.corpus,
          max.words = 100,
          random.order=FALSE, 
          rot.per=0.30, 
          use.r.layout=FALSE, 
          colors=brewer.pal(8, "Paired"))
```


# Geographical Analysis 


### HeatMap of Applied Loan Amount by State
```{r}
library(choroplethr)
library(choroplethrMaps)


df_approved$region <- df_approved$addr_state
df_approved$region <- as.factor(df_approved$region)
levels(df_approved$region)<- c("alaska", "alabama","arkansas", "arizona", "california","colorado","connecticut","district of columbia","delaware","florida","georgia","hawaii","iowa","idaho","illinois","indiana","kansas","kentucky","louisiana","massachusetts","maryland","maine","michigan","minnesota","missouri","mississippi","montana","north carolina","north dakota","nebraska","new hampshire","new jersey","new mexico","nevada","new york","ohio","oklahoma","oregon","pennsylvania","rhode island","south carolina","south dakota","tennessee","texas","utah","virginia","vermont","washington","wisconsin","west virginia","wyoming")

state_by_loan_amount <-
df_approved %>% group_by(region) %>%
  summarise(value = sum(loan_amnt, na.rm=TRUE))

state_choropleth(state_by_loan_amount, title = "Loan Amount by State", legend="loan_amnt")
```

```{r}
# df_approved$region <- df_approved$addr_state
# df_approved$region <- as.factor(df_approved$region)
# levels(df_approved$region)<- c("alaska", "alabama","arkansas", "arizona", "california","colorado","connecticut","district of columbia","delaware","florida","georgia","hawaii","iowa","idaho","illinois","indiana","kansas","kentucky","louisiana","massachusetts","maryland","maine","michigan","minnesota","missouri","mississippi","montana","north carolina","north dakota","nebraska","new hampshire","new jersey","new mexico","nevada","new york","ohio","oklahoma","oregon","pennsylvania","rhode island","south carolina","south dakota","tennessee","texas","utah","virginia","vermont","washington","wisconsin","west virginia","wyoming")

# all_states <- map_data("state")
# state_by_loan <-df_approved %>% group_by(region) %>% summarise(value = sum(loan_amnt, na.rm=TRUE))
# state_by_loan$region <- as.character(state_by_loan$region)

# Total <- merge(all_states, state_by_loan, by="region")

# p <- ggplot()
# p <- p + geom_polygon(data=Total, aes(x=long, y=lat, group = group, fill=Total$value),colour="white") + scale_fill_continuous(low = "skyblue", high = "darkblue", guide="colorbar")
# P1 <- p + theme_bw()  + labs(fill = "Gradient of loan amount" ,title = "Heat Map of loan amount in all states", x="", y="")
# P1 + scale_y_continuous(breaks=c()) + scale_x_continuous(breaks=c()) + theme(panel.border =  element_blank())
```

### HeatMap of Average Income by State
```{r}
state_by_avg_annual_inc <-
df_approved %>% group_by(region) %>%
  summarise(value = mean(annual_inc, na.rm=TRUE))

state_choropleth(state_by_avg_annual_inc, title = "Average Annual Income by State", legend="annual_inc", num_colors=8)

```

### Charged Off Loans Proportion by State
```{r}
options(repr.plot.width=10, repr.plot.height=10)
charged_off_df %>% group_by(addr_state) %>% 
summarise(charged_off.pct = n() * 100 / 204282) %>% ggplot(aes(x = reorder(addr_state, charged_off.pct), y = charged_off.pct)) +
geom_bar(stat = "identity", fill = "lightblue2", aes(color = I('white')), size = 0.1)+coord_flip()+xlab("States") + ylab("Percent of Charged Off Loans")
```

# Fully Paid Loans Proportion by State
```{r}
options(repr.plot.width=10, repr.plot.height=10)

fully_paid_df %>% group_by(addr_state) %>% 
summarise(fully_paid.pct = n() * 100 / 204282) %>% ggplot(aes(x = reorder(addr_state, fully_paid.pct), y = fully_paid.pct)) +
geom_bar(stat = "identity", fill = "lightblue2", aes(color = I('white')), size = 0.1)+coord_flip()+xlab("States") + ylab("Percent of Fully Paid Loans")
```

# Bivariate EDA

```{r}
# Creating new variable lstatus with only two categories: bad loans(late, charged off, default) and not bad loans

bad_indicators <- c("Charged Off ",
                    "Default",
                    "Does not meet the credit policy. Status:Charged Off",
                    "Default Receiver", 
                    "Late (16-30 days)",
                    "Late (31-120 days)")

# lstatus = 1 if loan is bad ; lstatus = 0 if loan is not bad
df_approved$lstatus <- ifelse(df_approved$loan_status %in% bad_indicators, 1,
                          ifelse(df_approved$loan_status=="", NA, 0)
                          )
```

### Requested Loan Amount v.s. Loan Status 

```{r}
ggplot(df_approved, aes(loan_status, loan_amnt)) + geom_boxplot(aes(fill = loan_status)) + theme(axis.text.x = element_blank()) +
  labs(list(
    title = "Loan amount by status",
    x = "Status",
    y = "Amount"))
```

### Requested Loan Amount v.s. Loan Status(Not Bad v.s. Bad)
```{r}
df_approved %>% ggplot(aes(as.factor(lstatus), loan_amnt)) + geom_boxplot(aes(fill=lstatus)) +
  labs(list(
    title = "Loan amount by Loan Status (Not Bad v.s. Bad Loans)",
    x = "Status",
    y = "Applied Amount"))
```

```{r}
# T-test: Do bad loans have higher applied loan amounts than not bad loans on average and is this statistically significant?
t.test(loan_amnt ~ lstatus, data=df_approved)
```

```{r}
# Independent 2-group Mann-Whitney U Test: : Do bad loans have higher applied loan amounts than not bad loans on average and is this statistically significant?
wilcox.test(loan_amnt ~ lstatus, data=df_approved)
```

### Average current balance of all accounts by Loan Staus(Not Bad v.s. Bad Loans)
```{r}
df_approved %>% ggplot(aes(as.factor(lstatus), avg_cur_bal)) + geom_boxplot(aes(fill=lstatus)) +
  labs(list(
    title = "Average current balance of all accounts by Loan Status (Not Bad v.s. Bad Loans)",
    x = "Status",
    y = "Average Current Balance of all Accounts"))
```
```{r}
# Is the difference in Average current balance of all accounts between bad loans and not bad loans statistically significant? (parametric test)
t.test(avg_cur_bal ~ lstatus, data=df_approved)
```

```{r}
# Is the difference in Average current balance of all accounts between bad loans and not bad loans statistically significant? (non-parametric test)
wilcox.test(avg_cur_bal ~ lstatus, data=df_approved)
```

### Interest Rate by Grade

```{r}
# Taking out the % sign at the end of each interest rate and changing datatype from string to float
df_approved$int_rate <- substr(df_approved$int_rate,1, nchar(df_approved$int_rate)-1)
df_approved$int_rate <- as.numeric(df_approved$int_rate)
```

```{r}
ggplot(df_approved , aes(x = grade , y = int_rate , fill = grade)) + 
        geom_boxplot() + 
        labs(y = 'Interest Rate' , x = 'Grade')
```

```{r}
# Independent 2-group Mann-Whitney U Test on loan amount difference between grade A applicants and grade B applicants
wilcox.test(int_rate ~ grade, data=df_approved%>%filter(grade=='A' | grade=='B'))
```

```{r}
# Independent 2-group Mann-Whitney U Test on loan amount difference between grade B applicants and grade C applicants
wilcox.test(int_rate ~ grade, data=df_approved%>%filter(grade=='B' | grade=='C'))
```

```{r}
# Independent 2-group Mann-Whitney U Test on loan amount difference between grade B applicants and grade C applicants
wilcox.test(int_rate ~ grade, data=df_approved%>%filter(grade=='C' | grade=='D'))
```

```{r}
# Independent 2-group Mann-Whitney U Test on loan amount difference between grade D applicants and grade E applicants
wilcox.test(int_rate ~ grade, data=df_approved%>%filter(grade=='D' | grade=='E'))
```

```{r}
# Independent 2-group Mann-Whitney U Test on loan amount difference between grade E applicants and grade F applicants
wilcox.test(int_rate ~ grade, data=df_approved%>%filter(grade=='E' | grade=='F'))
```

```{r}
# Independent 2-group Mann-Whitney U Test on loan amount difference between grade F applicants and grade G applicants
wilcox.test(int_rate ~ grade, data=df_approved%>%filter(grade=='F' | grade=='G'))
```

### Applied Loan Amount v.s. Home Ownership
```{r}
ggplot(data=df_approved, aes(home_ownership,loan_amnt,fill=home_ownership))+geom_boxplot(outlier.color = "blue")+labs(title="Box Plot of Loan Amount by Homeownership")
```

```{r}
# Is the difference in loan amount between home owners and those who bought their houses through mortgage statistically significant? (parametric test)
t.test(loan_amnt ~ home_ownership, data=df_approved%>%filter(home_ownership =='MORTGAGE' | home_ownership=='OWN'))
```

```{r}
# Is the difference in loan amount between home owners and those who bought their houses through mortgage statistically significant? (non-parametric test)
wilcox.test(loan_amnt ~ home_ownership, data=df_approved%>%filter(home_ownership =='MORTGAGE' | home_ownership=='OWN'))
```

```{r}
# Is the difference in loan amount between home owners and those who rented their housesstatistically significant? (parametric test)
t.test(loan_amnt ~ home_ownership, data=df_approved%>%filter(home_ownership =='RENT' | home_ownership=='OWN'))
```

```{r}
# Is the difference in loan amount between home owners and those who rented their housesstatistically significant? (non-parametric test)
wilcox.test(loan_amnt ~ home_ownership, data=df_approved%>%filter(home_ownership =='RENT' | home_ownership=='OWN'))
```

### Loan Status Proportion by Grade

```{r}
library(vcd)

mosaic(~lstatus+grade, 
       data=df_approved, 
       gp=shading_max,
       split_vertical=T)
```

### Loan Status Proportions by Employment Length
```{r}
ggplot(filter(df_approved, emp_length != 'n/a'), aes(x =emp_length, fill = loan_status)) + 
geom_bar(stat='count', position='fill', aes(color = I('black')), size = 0.1) +labs(x = 'emp_length') + 
  scale_fill_discrete(name="Loan_Status") + 
theme(axis.text.x = element_text(angle = 45, vjust = 0.8, hjust=1))
```

### Purpose of Loans v.s. Percentage of Charged Off Loans
```{r}
charged_off_df %>% group_by(purpose) %>% summarise(charged_off.pct = n() * 100 / 204282) %>% 
ggplot(aes(x = reorder(purpose, charged_off.pct), y = charged_off.pct)) + geom_bar(stat = "identity", fill =  "coral", aes(color = I('black')), size = 0.1) + coord_flip() + xlab("Purpose") + ylab("Percent of Charged Off Loans")
```

### Purpose of Loans v.s. Percentage of Fully Paid Loans
```{r}
fully_paid_df %>% group_by(purpose) %>% summarise(fully_paid.pct = n() * 100 / 204282) %>% 
ggplot(aes(x = reorder(purpose, fully_paid.pct), y = fully_paid.pct)) + geom_bar(stat = "identity", fill =  "coral", aes(color = I('black')), size = 0.1) + coord_flip() + xlab("Purpose") + ylab("Percent of Fully Paid Back Loans")
```

### Applied Loan Amount v.s. Annual Income
```{r}
ggplot(df_approved[sample(244179 , 10000) , ], aes(x = annual_inc , y = loan_amnt , color = int_rate)) +
        geom_point(alpha = 0.5 , size = 1.5) + 
        geom_smooth(se = F , color = 'darkred' , method = 'loess') +
        xlim(c(0 , 300000)) + 
        labs(x = 'Annual Income' , y = 'Applied Loan Ammount' , color = 'Interest Rate')
```

```{r}
# Linear Regression on Applied Loan Amount v.s. Annual Income
summary(lm(loan_amnt~annual_inc, data=df_approved))
```

### Applied Loan Amount v.s. Purpose of Loan
```{r}
df_approved %>%select(purpose,loan_amnt)%>%ggplot(aes(x=purpose,y=loan_amnt,col=purpose))+geom_jitter(size=0.5)+labs(title="Applied Loan Amounts by Purpose", las=2)
```


### Average Interest Rate for Loans
```{r}
df_approved %>%select(issue_d_year,grade,int_rate)%>%group_by(issue_d_year,grade)%>%summarise(avg_intrate=mean(int_rate))%>%
  ggplot(aes(x=issue_d_year,y=avg_intrate,group=grade,col=grade))+geom_line(size=1.5)+labs(x="Year",y="Avg Interest Rate",title="Average Interest Rate for Loans")
```




















