---
title: "Data Wrangling and Cleaning"
author: "Seungjun (Josh) Kim"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### Library and Data Loading

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Load in Libraries / Packages

library(tidyverse)
library(ggthemes)
library(GGally)
library(ggExtra)
library(caret)
library(glmnet)
library(corrplot)
library(leaflet)
library(kableExtra)
library(RColorBrewer)
library(plotly)
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

set.seed(42)

# Read in 2019 NYC airbnb Data
abnb19 <- read.csv("AB_NYC_2019.csv", encoding="UTF-8", stringsAsFactors = F, na.strings=c(""))

# Read in 2020 NYC Airbnb Data
abnb20 <- read.csv("listings.csv", encoding="UTF-8", stringsAsFactors = F, na.strings=c(""))

# Read in tract-level ACS 2017 Data
acs17 <- read.csv("acs2017_census_tract_data.csv")

nyc_tract <- read.csv("nyc_census_tracts.csv")

# Read in 2016,2017,and 2018 NYC airbnb Listings Data
abnb16 <- read.csv("listings16.csv", encoding="UTF-8", stringsAsFactors = F, na.strings=c(""))
abnb17 <- read.csv("listings17.csv", encoding="UTF-8", stringsAsFactors = F, na.strings=c(""))
abnb18 <- read.csv("listings18.csv", encoding="UTF-8", stringsAsFactors = F, na.strings=c(""))
```

### Taking a peak into Data

```{r}
# First five lines of 2020 nyc airbnb data
head(abnb20)
```

```{r}
# First five lines of 2019 nyc airbnb data
head(abnb19)
```

### Remove some unncessary variables

```{r}
# Remove unnecessary columns
vars_delete <- c("id", "host_id", "host_name")
abnb19 <- select(abnb19, -vars_delete)
abnb20 <- select(abnb20, -vars_delete)
abnb16 <- select(abnb16, -vars_delete)
abnb17 <- select(abnb17, -vars_delete)
abnb18 <- select(abnb18, -vars_delete)
```

### Fixing Data Types of Features

```{r}
# Change datatype of factor features from char to fct
fact_feat <- c("neighbourhood_group", "neighbourhood", "room_type")
abnb16[fact_feat] <- map(abnb16[fact_feat], as.factor)
abnb17[fact_feat] <- map(abnb17[fact_feat], as.factor)
abnb18[fact_feat] <- map(abnb18[fact_feat], as.factor)
abnb19[fact_feat] <- map(abnb19[fact_feat], as.factor)
abnb20[fact_feat] <- map(abnb20[fact_feat], as.factor)
```

```{r}
# Change datatype of numerical features from char to dbl

Sys.setlocale("LC_ALL", "C")

dbl_feat <- c("latitude","longitude","price","minimum_nights","number_of_reviews","reviews_per_month","calculated_host_listings_count","availability_365")
abnb16[dbl_feat] <- map(abnb16[dbl_feat], as.double)
abnb17[dbl_feat] <- map(abnb17[dbl_feat], as.double)
abnb18[dbl_feat] <- map(abnb18[dbl_feat], as.double)
abnb19[dbl_feat] <- map(abnb19[dbl_feat], as.double)
abnb20[dbl_feat] <- map(abnb20[dbl_feat], as.double)
```
### Missing Values

```{r}
# Checking how many missing values are there in each variable for year 2019
sapply(abnb19, function(x) sum(is.na(x)))
```

```{r}
# Checking how many missing values are there in each variable for year 2020
sapply(abnb20, function(x) sum(is.na(x)))
```

```{r}
# Checking how many missing values are there in each variable for year 2016
sapply(abnb16, function(x) sum(is.na(x)))
```

```{r}
# Checking how many missing values are there in each variable for year 2017
sapply(abnb17, function(x) sum(is.na(x)))
```

```{r}
# Checking how many missing values are there in each variable for year 2017
sapply(abnb18, function(x) sum(is.na(x)))
```

```{r}
# Replace NAs with 0 or appropriate value

abnb16$reviews_per_month <- ifelse(is.na(abnb16$reviews_per_month), 0, abnb16$reviews_per_month)
abnb17$reviews_per_month <- ifelse(is.na(abnb17$reviews_per_month), 0, abnb17$reviews_per_month)
abnb18$reviews_per_month <- ifelse(is.na(abnb18$reviews_per_month), 0, abnb18$reviews_per_month)
abnb19$reviews_per_month <- ifelse(is.na(abnb19$reviews_per_month), 0, abnb19$reviews_per_month)
abnb20$reviews_per_month <- ifelse(is.na(abnb20$reviews_per_month), 0, abnb20$reviews_per_month)

abnb16$number_of_reviews <- ifelse(is.na(abnb16$number_of_reviews), 0, abnb16$number_of_reviews)
abnb17$number_of_reviews <- ifelse(is.na(abnb17$number_of_reviews), 0, abnb17$number_of_reviews)
abnb18$number_of_reviews <- ifelse(is.na(abnb18$number_of_reviews), 0, abnb18$number_of_reviews)
abnb19$number_of_reviews <- ifelse(is.na(abnb19$number_of_reviews), 0, abnb19$number_of_reviews)
abnb20$number_of_reviews <- ifelse(is.na(abnb20$number_of_reviews), 0, abnb20$number_of_reviews)

abnb16$last_review <- ifelse(is.na(abnb16$last_review), "No Review", abnb16$last_review)
abnb17$last_review <- ifelse(is.na(abnb17$last_review), "No Review", abnb17$last_review)
abnb18$last_review <- ifelse(is.na(abnb18$last_review), "No Review", abnb18$last_review)
abnb19$last_review <- ifelse(is.na(abnb19$last_review), "No Review", abnb19$last_review)
abnb20$last_review <- ifelse(is.na(abnb20$last_review), "No Review", abnb20$last_review)
```

```{r}
# Omit observations that has any missing value in any of the columns

abnb16 <- na.omit(abnb16)
abnb17 <- na.omit(abnb17)
abnb18 <- na.omit(abnb18)
abnb19 <- na.omit(abnb19)
abnb20 <- na.omit(abnb20)
```

### Checking if there are any anomalies in "Price" variable

Price is going to be our "response" variable for prediction later, so it's important to identify any weird values (e.g. minus or 0)

```{r}
# Number of observations that have price 0 or lower in 2016 data
dim(abnb16 %>% filter(price <= 0))[1]
```

```{r}
# Number of observations that have price 0 or lower in 2017 data
dim(abnb17 %>% filter(price <= 0))[1]
```

```{r}
# Number of observations that have price 0 or lower in 2018 data
dim(abnb18 %>% filter(price <= 0))[1]
```

```{r}
# Number of observations that have price 0 or lower in 2019 data
dim(abnb19 %>% filter(price <= 0))[1]
```

```{r}
# Number of observations that have price 0 or lower in 2020 data

dim(abnb20 %>% filter(price <= 0))[1]
```

```{r}
# We Drop observations that have price = 0 or lower because this may be dummy / example entries on the website or some indication of an error in the system which shouldn't be included as part of the analysis

abnb18 <- abnb18 %>% filter(price > 0)
abnb20 <- abnb20 %>% filter(price > 0)
```

### Merge demographic, socioeconomic variables from ACS data into Airbnb data

```{r}
# Add "borough" information from nyc_tract data to ACS17 data
acs <- acs17 %>% inner_join(nyc_tract %>% select('CensusTract', 'Borough'), by=c('TractId'='CensusTract'))
```

```{r}
# Average values of variables of tracts for each borough

acs_avg_borough <- acs %>% group_by(Borough) %>% summarize(totalpop = sum(TotalPop, na.rm=TRUE), men_avg = mean(Men, na.rm=TRUE), wom_avg = mean(Women, na.rm=TRUE),
                                        hisp_avg = mean(Hispanic, na.rm=TRUE), white_avg = mean(White, na.rm=TRUE), black_avg = mean(Black, na.rm=TRUE), 
                                        native_avg = mean(Native, na.rm=TRUE), asian_avg = mean(Asian, na.rm=TRUE), pacific_avg = mean(Pacific, na.rm=TRUE),
                                        inc_avg = mean(Income, na.rm=TRUE), pov_avg = mean(Poverty, na.rm=TRUE), unemp_avg = mean(Unemployment, na.rm=TRUE),
                                        childpov_avg = mean(ChildPoverty, na.rm=TRUE), selfemp_avg = mean(SelfEmployed, na.rm=TRUE))
```

```{r}
abnb16 <- abnb16 %>% left_join(acs_avg_borough, by=c('neighbourhood_group' = 'Borough'))
abnb17 <- abnb17 %>% left_join(acs_avg_borough, by=c('neighbourhood_group' = 'Borough'))
abnb18 <- abnb18 %>% left_join(acs_avg_borough, by=c('neighbourhood_group' = 'Borough'))
abnb19 <- abnb19 %>% left_join(acs_avg_borough, by=c('neighbourhood_group' = 'Borough'))
abnb20 <- abnb20 %>% left_join(acs_avg_borough, by=c('neighbourhood_group' = 'Borough'))
```


```{r}
# Save these wrangled CSV files

# write.csv(abnb16,'airbnb16.csv')
# write.csv(abnb17,'airbnb17.csv')
# write.csv(abnb18,'airbnb18.csv')
# write.csv(abnb19,'airbnb19.csv')
# write.csv(abnb20,'airbnb20.csv')
```


