---
title: "Some Explanatory Data Analysis (EDA)"
output: html_notebook
author: "Seungjun (Josh) Kim"
---

### Import Libraries and Data

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(ggthemes)
library(GGally)
library(ggExtra)
library(caret)
library(glmnet)
library(corrplot)
library(leaflet)
library(kableExtra)
library(RColorBrewer)
library(plotly)
library(tm)
library(lubridate)
library(caret)
library(lattice)
# library(wordcloud)
# library(MASS)
# library(NLP)
```

```{r}
# Read in Wrangled Data

abnb16 <- read.csv("airbnb16.csv")
abnb17 <- read.csv("airbnb17.csv")
abnb18 <- read.csv("airbnb18.csv")
abnb19 <- read.csv("airbnb19.csv")
abnb20 <- read.csv("airbnb20.csv")
```

```{r}
# Drop Unnecessary columns (we don't need the first column which is the index column)

abnb16 <- dplyr::select(abnb16,-1)
abnb17 <- dplyr::select(abnb17,-1)
abnb18 <- dplyr::select(abnb18,-1)
abnb19 <- dplyr::select(abnb19,-1)
abnb20 <- dplyr::select(abnb20,-1)
```

```{r}
# Merged all datasets into one

abnb16$year <- "2016" 
abnb17$year <- "2017"
abnb18$year <- "2018"
abnb19$year <- "2019" 
abnb20$year <- "2020"

abnb <- rbind(abnb16, abnb17,abnb18, abnb19, abnb20)

abnb$year <- as.factor(abnb$year)
```

```{r}
th <- theme_fivethirtyeight() + theme(axis.title = element_text(), axis.title.x = element_text())
```

### Price Distribution by room type

```{r}
# Boxplots of price distributions by room type
ggplot(abnb, aes(x = room_type, y = price)) +
  geom_boxplot(aes(fill = room_type)) + 
  th + 
  xlab("Room type") + 
  ylab("Price") + scale_y_log10() +
  ggtitle("Price Distributions by room type in 2016-2020") + 
  geom_hline(yintercept = mean(abnb$price), color = "purple", linetype = 2) + facet_grid(~year) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

```{r}
# Mean and Meidan price by room type over time
abnb %>% group_by(year, room_type) %>% summarize(median = median(price), mean = mean(price)) %>% arrange(room_type,year)
```

Previous literature showed that entire homes on the platform of Airbnb in New York City cost on average of twice as much as private rooms; however, the difference between the price of a private room and a shared room is not as significant at approximately $15. How do these price differences look like in 2019 and 2020? Are they statistically significant?

##### Average price of entire apartment v.s. private room
```{r}
# Regression of price on entire apartment v.s. private room in 2017
summary(lm(price ~ room_type, data = abnb %>% filter(year=="2017" & room_type %in% c('Entire home/apt','Private room'))))
```

```{r}
# Regression of price on entire apartment v.s. private room in 2020
summary(lm(price ~ room_type, data = abnb %>% filter(year=="2020" & room_type %in% c('Entire home/apt','Private room'))))
```

```{r}
# controlling for other variables in 2020
summary(lm(price ~ room_type + neighbourhood_group + minimum_nights + number_of_reviews + availability_365 + calculated_host_listings_count,
           data = abnb %>% filter(year=="2020" & room_type %in% c('Private room','Entire home/apt'))))
```

##### Average price of private room v.s. shared room

```{r}
# Regression of price on private room v.s. shared room in 2017
summary(lm(price ~ room_type, data = abnb %>% filter(year=="2017" & room_type %in% c('Private room','Shared room'))))
```

```{r}
# Regression of price on private room v.s. shared room in 2020
summary(lm(price ~ room_type, data = abnb %>% filter(year=="2020" & room_type %in% c('Private room','Shared room'))))
```

```{r}
# controlling for other variables in 2020
summary(lm(price ~ room_type + neighbourhood_group + minimum_nights + number_of_reviews + availability_365 + calculated_host_listings_count,
           data = abnb %>% filter(year=="2020" & room_type %in% c('Private room','Shared room'))))
```



### Price Distribution by Neighborhood Group

```{r}
ggplot(abnb, aes(x = neighbourhood_group, y = price)) +
  geom_boxplot(aes(fill = neighbourhood_group)) + 
  th + ylab("Price") + scale_y_log10() +
  ggtitle("Price Distributions by Neighborhood Group in 16'-20'") + 
  geom_hline(yintercept = mean(abnb$price), color = "purple", linetype = 2) + facet_grid(~year) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

```{r}
abnb %>% group_by(year, neighbourhood_group) %>% summarize(median = median(price), mean = mean(price)) %>% arrange(neighbourhood_group,year)
```

Previous literature showed that, in 2017, the average Airbnb accommodation prices are more expensive in Manhattan and Brooklyn. Does this hold true in 2020 as well?

##### Average price in Manhattan v.s. Queens in 2017, 2019 and 2020

```{r}
# 2017
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2017" & neighbourhood_group %in% c('Manhattan','Queens'))))
```

```{r}
# 2019
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2019" & neighbourhood_group %in% c('Manhattan','Queens'))))
```

```{r}
# 2020
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2020" & neighbourhood_group %in% c('Manhattan','Queens'))))
```

```{r}
# controlling for other variables in 2020
summary(lm(price ~ neighbourhood_group + minimum_nights + number_of_reviews + availability_365 + calculated_host_listings_count,
           data = abnb %>% filter(year=="2020" & neighbourhood_group %in% c('Manhattan','Queens'))))
```

##### Average price in Brooklyn v.s. Queens in 2017, 2019 and 2020

```{r}
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2017" & neighbourhood_group %in% c('Queens','Brooklyn'))))
```

```{r}
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2019" & neighbourhood_group %in% c('Queens','Brooklyn'))))
```

```{r}
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2020" & neighbourhood_group %in% c('Queens','Brooklyn'))))
```

```{r}
# controlling for other variables in 2020
summary(lm(price ~ neighbourhood_group + minimum_nights + number_of_reviews + availability_365 + calculated_host_listings_count,
           data = abnb %>% filter(year=="2020" & neighbourhood_group %in% c('Brooklyn','Queens'))))
```

##### Average price in Manhattan v.s. Brooklyn in 2017, 2019 and 2020

```{r}
# 2017
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2017" & neighbourhood_group %in% c('Manhattan','Brooklyn'))))
```

```{r}
# 2019
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2019" & neighbourhood_group %in% c('Manhattan','Brooklyn'))))
```

```{r}
# 2020
summary(lm(price ~ neighbourhood_group, data = abnb %>% filter(year=="2020" & neighbourhood_group %in% c('Manhattan','Brooklyn'))))
```

```{r}
# controlling for other variables in 2020
summary(lm(price ~ neighbourhood_group + minimum_nights + number_of_reviews + availability_365 + calculated_host_listings_count,
           data = abnb %>% filter(year=="2020" & neighbourhood_group %in% c('Brooklyn','Manhattan'))))
```

### Number of listings by Room Type

```{r}
ggplot(abnb, aes(x=room_type, y=(..count..), fill=room_type)) + geom_bar() + facet_grid(~year) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +ggtitle("Number of Listings Breakdown by Room Type in 2016-2020") + ylab("Number of Listings")
```

### Number of listings by neighbourhood area in 2016-2020

```{r}
abnb %>% group_by(year, neighbourhood_group) %>% tally() %>% 
  ggplot(aes(x = reorder(neighbourhood_group, n), n)) +
  geom_bar(stat = "identity", fill = "orange") +
  theme_fivethirtyeight() +
  ggtitle("Number of listings by neighbourhood group 16'-20'") +
  geom_text(aes(x = neighbourhood_group, y = 1, label = paste0(n)), hjust=-1.5, vjust=.5, size = 4, fontface = 'bold') +
  coord_flip() + facet_grid(~year)
```

```{r}
listng_n_by_neighb_df <- rbind(abnb %>% group_by(year, neighbourhood_group) %>% tally(), abnb %>% group_by(year, neighbourhood_group) %>% tally())
```

```{r}
listng_n_by_neighb_df %>% filter(neighbourhood_group %in% c('Manhattan', 'Brooklyn')) %>%
  ggplot(aes(x = year, y = n, group=neighbourhood_group, color=neighbourhood_group)) +
  geom_line() + 
  theme_fivethirtyeight() +
  ggtitle("Number of listings in Manhattan and Brooklyn in 16'-20'")
```

```{r}
listng_n_by_neighb_df %>% filter(neighbourhood_group %in% c('Bronx', 'Queens', 'Staten Island')) %>%
  ggplot(aes(x = year, y = n, group=neighbourhood_group, color=neighbourhood_group)) +
  geom_line() + 
  theme_fivethirtyeight() +
  ggtitle("Number of listings in Queens,Bronx,and SI in 16'-20'")
```

```{r}
# 2017 - Leaflet Map of Listings

 palett17 <- colorFactor(palette = c("red", "green", "blue", "purple", "yellow"), domain = abnb17$neighbourhood_group)
 
 leaflet(data = abnb17) %>% addProviderTiles(providers$CartoDB.DarkMatterNoLabels) %>%  addCircleMarkers(~longitude, ~latitude, color = ~palett17(neighbourhood_group), weight = 1, radius=1, fillOpacity = 0.1, opacity = 0.1,
                                                                                                        label = paste("Name:", abnb17$name)) %>% 
     addLegend("bottomright", pal = palett17, values = ~neighbourhood_group,
     title = "Neighbourhood groups",
     opacity = 1
   )
```

```{r}
# 2020 - Leaflet Map of Listings

 palett20 <- colorFactor(palette = c("red", "green", "blue", "purple", "yellow"), domain = abnb20$neighbourhood_group)
 
 leaflet(data = abnb20) %>% addProviderTiles(providers$CartoDB.DarkMatterNoLabels) %>%  addCircleMarkers(~longitude, ~latitude, color = ~palett20(neighbourhood_group), weight = 1, radius=1, fillOpacity = 0.1, opacity = 0.1,
                                                                                                        label = paste("Name:", abnb20$name)) %>% 
     addLegend("bottomright", pal = palett20, values = ~neighbourhood_group,
     title = "Neighbourhood groups",
     opacity = 1
   )
```

### Number of Reviews v.s. Price

```{r}
ggplot(abnb %>% filter(year %in% c("2017","2020") & neighbourhood_group %in% c('Queens','Bronx','Staten Island')), 
       aes(x=number_of_reviews, y=price, color=neighbourhood_group, alpha=0.9)) + 
  geom_point() + facet_grid(~year) + ggtitle("# of Reviews v.s. Price of Listings in Queens, Bronx, and Staten Island in 17',20'")
```

```{r}
ggplot(abnb %>% filter(year %in% c("2017","2020") & neighbourhood_group %in% c('Manhattan','Brooklyn')), 
       aes(x=number_of_reviews, y=price, color=neighbourhood_group, alpha=0.1)) + 
  geom_point() + facet_grid(~year) + ggtitle("# of Reviews v.s. Price of Listings in Manhattan and Brooklyn in 17',20'")
```





